---
title: "Getting Started with BBMF"
author: "Adolphus Wagala"

format:
  pdf:
    toc: true
    number-sections: true
    keep-tex: false

vignette: >
  %\VignetteIndexEntry{Getting Started with BBMF}
  %\VignetteEngine{quarto::pdf}
  %\VignetteEncoding{UTF-8}

execute:
  echo: true
  warning: false
  message: false

references:
  - id: wagala2026bbmf
    type: article-journal
    title: "A Bayesian Boolean Matrix Factorization with Application to Copy Number Analysis in Cancer"
    author:
      - family: Wagala
        given: Adolphus
      - family: Samur
        given: Mehmet
      - family: Parmigiani
        given: Giovanni
    issued:
      - year: 2026
    URL: "https://arxiv.org/abs/2606.17491"
    note: "arXiv:2606.17491"
---

## Introduction

The `BBMF` package implements the Bayesian Boolean matrix factorization methods presented in *A Bayesian Boolean Matrix Factorization with Application to Copy Number Analysis in Cancer* by A. Wagala, M. Samur, and G. Parmigiani [@wagala2026bbmf].

The package is designed for binary matrix data. For a binary matrix $X$ of dimension $K \times G$, BBMF represents

$$X \approx W \circ H,$$

where $W$ is a $K \times R$ binary sample-factor matrix, $H$ is an $R \times G$ binary factor-feature matrix, and $\circ$ denotes Boolean matrix multiplication.

The package provides tools for fitting sparse Bayesian Boolean matrix factorizations, posterior inference, Boolean reconstruction, factor alignment, latent-factor recovery assessment, reconstruction diagnostics, posterior uncertainty quantification, and visualization.

This vignette illustrates a basic BBMF workflow:

- simulate a binary matrix;
- initialize a Boolean factorization;
- fit sparse BBMF;
- inspect MCMC diagnostics;
- select a representative posterior solution;
- visualize the estimated latent factors;
- align estimated factors;
- assess latent-factor recovery using AFAS;
- evaluate Boolean reconstruction;
- summarize posterior uncertainty in $X$ and $H$.

## Software requirements

`BBMF` builds on several existing R packages.

In particular, the package uses `rBMF` for the Asso Boolean matrix factorization used for initialization and comparison. Other packages provide supporting computational and visualization tools. For example, this vignette uses `reshape2` for data preparation and `patchwork` for arranging plots, while the plotting functions in `BBMF` use `ggplot2`.

Before reproducing this vignette, users should ensure that the supporting packages are installed.

```{r}
#| label: install-packages
#| eval: false

install.packages(
  c(
    "rBMF",
    "ggplot2",
    "reshape2",
    "patchwork",
    "clue",
    "abind",
    "pbmcapply",
    "progress"
  )
)
```

When `BBMF` is installed as an R package, packages listed as required dependencies are normally installed automatically.

## Load BBMF

```{r}
#| label: setup

library(BBMF)

set.seed(101)
```

## Simulate a binary matrix

We generate a small example with 64 samples, 44 genomic features, and four latent factors.

```{r}
#| label: simulation-setup

K <- 64
G <- 44
R <- 4

arms <- unlist(
  lapply(
    1:22,
    function(i) paste0("chr", i, c("p", "q"))
  )
)

sample_ids <- paste0(
  "Sample_",
  seq_len(K)
)

factor_names <- paste0(
  "Factor_",
  seq_len(R)
)
```

### Generate the true factor-feature matrix

We first generate a binary factor-feature matrix $H$ containing four latent feature patterns.

```{r}
#| label: simulate-H

theta_high <- c(
  0.80,
  0.45,
  0.65,
  0.05
)

theta_low <- c(
  0.05,
  0.05,
  0.25,
  0.05
)

regions <- list(
  1:22,
  20:40,
  25:44,
  1:44
)

H_true <- matrix(
  0,
  nrow = R,
  ncol = G
)

for (r in seq_len(R)) {

  for (g in seq_len(G)) {

    if (g %in% regions[[r]]) {

      H_true[r, g] <- stats::rbinom(
        1,
        1,
        theta_high[r]
      )

    } else {

      H_true[r, g] <- stats::rbinom(
        1,
        1,
        theta_low[r]
      )
    }
  }
}

rownames(H_true) <- factor_names
colnames(H_true) <- arms
```

### Visualize the true factor-feature matrix

The function `matrixPlot()` can be used to visualize binary matrices.

```{r}
#| label: plot-true-H
#| fig-width: 11
#| fig-height: 3

H_true_df <- reshape2::melt(
  H_true
)

H_true_df$Var1 <- factor(
  H_true_df$Var1,
  levels = factor_names
)

pH <- matrixPlot(
  H_true_df,
  Var2,
  Var1,
  value,
  title = "True Factor-Feature Matrix H"
)

pH
```

### Generate the true sample-factor matrix

Factor prevalences are generated from a Beta distribution and used to generate the sample-factor matrix $W$.

```{r}
#| label: simulate-W

nu <- stats::rbeta(
  R,
  shape1 = 2,
  shape2 = 6
)

W_true <- matrix(
  0,
  nrow = K,
  ncol = R
)

for (r in seq_len(R)) {

  W_true[, r] <- stats::rbinom(
    K,
    size = 1,
    prob = nu[r]
  )
}

rownames(W_true) <- sample_ids
colnames(W_true) <- factor_names
```

### Visualize the true sample-factor matrix

```{r}
#| label: plot-true-W
#| fig-width: 5
#| fig-height: 8

W_true_df <- reshape2::melt(
  W_true
)

W_true_df$Var2 <- factor(
  W_true_df$Var2,
  levels = factor_names
)

pW <- matrixPlot(
  W_true_df,
  Var2,
  Var1,
  value,
  title = "True Sample-Factor Matrix W"
)

pW
```

### Construct the true Boolean matrix

The function `bool_prod()` computes the Boolean product of $W$ and $H$.

```{r}
#| label: simulate-X

X_true <- bool_prod(
  W_true,
  H_true
)

rownames(X_true) <- sample_ids
colnames(X_true) <- arms
```

### Add noise

The function `noise_bool()` can be used to randomly flip entries of a binary matrix. Here, 20% of the entries are flipped to generate the observed matrix.

```{r}
#| label: add-noise

X_noisy <- noise_bool(
  X_true,
  noise_level = 0.20,
  seed = 123
)

rownames(X_noisy) <- sample_ids
colnames(X_noisy) <- arms
```

### Visualize the true and observed matrices

```{r}
#| label: plot-X
#| fig-width: 12
#| fig-height: 7

X_true_df <- reshape2::melt(
  X_true
)

X_noisy_df <- reshape2::melt(
  X_noisy
)

pX <- matrixPlot(
  X_true_df,
  Var2,
  Var1,
  value,
  title = "True X"
)

pX_noisy <- matrixPlot(
  X_noisy_df,
  Var2,
  Var1,
  value,
  title = "Observed X"
)

patchwork::wrap_plots(
  pX,
  pX_noisy,
  ncol = 2
)
```

## Obtain an initial factorization

The function `Asso_BBMF()` provides an interface to the Asso Boolean matrix factorization implemented through the `rBMF` package.

The resulting factorization can be used to initialize BBMF.

```{r}
#| label: asso-initialization
#| results: hide

asso_fit <- Asso_BBMF(
  X = X_noisy,
  R = R,
  seed = 101
)

W_asso <- asso_fit$W_asso
H_asso <- asso_fit$H_asso
X_asso <- asso_fit$X_asso

rownames(W_asso) <- sample_ids
colnames(W_asso) <- factor_names

rownames(H_asso) <- factor_names
colnames(H_asso) <- arms

rownames(X_asso) <- sample_ids
colnames(X_asso) <- arms
```

### Visualize the Asso factor-feature matrix

```{r}
#| label: plot-asso-H
#| fig-width: 11
#| fig-height: 3

H_asso_df <- reshape2::melt(
  H_asso
)

H_asso_df$Var1 <- factor(
  H_asso_df$Var1,
  levels = factor_names
)

matrixPlot(
  H_asso_df,
  Var2,
  Var1,
  value,
  title = "Asso Estimate of H"
)
```

## Fit sparse BBMF

The main sparse BBMF fitting functions are `run_sparseBBMF()` for a single chain and `run_multiple_sparseBBMF()` for multiple chains.

For this introductory example, we use a small number of iterations and two chains so that the vignette can be reproduced quickly. Applications should generally use longer chains and perform appropriate convergence assessment.

```{r}
#| label: fit-bbmf
#| results: hide

n_iter <- 200
burn_in <- 0
n_chains <- 2

set.seed(2026)

psi_o <- stats::rbinom(
  G,
  1,
  0.5
)

fit_BBMF <- run_multiple_sparseBBMF(
  X = X_noisy,
  W_o = W_asso,
  H_o = H_asso,
  psi_o = psi_o,
  n_iter = n_iter,
  burn_in = burn_in,
  n_chains = n_chains,
  a1 = 1,
  a2 = 1,
  b1 = 1,
  b2 = 1,
  c1 = 1,
  c2 = 1,
  d1 = 1,
  d2 = 1
)
```

The returned object contains posterior samples of the latent matrices and model parameters for each MCMC chain.

## MCMC diagnostics

The function `diagnosBBMF()` calculates quantities such as the log-likelihood and log-posterior across MCMC samples.

```{r}
#| label: bbmf-diagnostics
#| results: hide

nn <- n_iter + 1

diagnostic_BBMF <- diagnosBBMF(
  X = X_noisy,
  BBMFresult = fit_BBMF,
  n_iter = nn,
  burn_in = 0,
  num_chains = n_chains
)

grouped_diagnostics <- group_diagnost_List(
  diagnostic_BBMF
)
```

For example, the sampled log-posterior values can be inspected using

```{r}
#| label: inspect-log-posterior

head(
  grouped_diagnostics$log_postr
)
```

## Select a representative posterior solution

For illustration, we select the sampled state with the largest log-posterior value across the chains.

```{r}
#| label: select-best-sample

logPost <- grouped_diagnostics$log_postr

logPost_mcmc <- logPost[
  -1,
  ,
  drop = FALSE
]

max_logPost <- max(
  as.matrix(logPost_mcmc),
  na.rm = TRUE
)

best_position <- which(
  as.matrix(logPost_mcmc) == max_logPost,
  arr.ind = TRUE
)[1, ]

best_iteration <- best_position[1]
best_chain <- best_position[2]

best_sample_index <- best_iteration + 1

best_fit <- fit_BBMF[[best_chain]]

W_max <- best_fit$samples_W[
  ,
  ,
  best_sample_index
]

H_max <- best_fit$samples_H[
  ,
  ,
  best_sample_index
]

rownames(W_max) <- sample_ids
colnames(W_max) <- factor_names

rownames(H_max) <- factor_names
colnames(H_max) <- arms

X_tilde_BBMF <- bool_prod(
  W_max,
  H_max
)

rownames(X_tilde_BBMF) <- sample_ids
colnames(X_tilde_BBMF) <- arms

cat(
  "Selected chain:", best_chain, "\n",
  "Iteration:", best_iteration, "\n",
  "Log-posterior:", round(max_logPost, 2), "\n"
)
```

## Factor similarity and alignment

The labels assigned to latent factors are arbitrary. Consequently, estimated factors should be aligned before comparing them with a reference factorization.

`factor_similarity()` calculates Jaccard similarity between factors and uses the Hungarian algorithm to identify the optimal alignment.

```{r}
#| label: factor-similarity

sim_BBMF <- factor_similarity(
  H_ref = H_true,
  H_est = H_max
)

sim_BBMF$alignment$matching

sim_BBMF$mean_similarity
```

The aligned similarity matrix can be visualized using `plot_factor_similarity()`.

```{r}
#| label: plot-factor-similarity
#| fig-width: 6
#| fig-height: 5

plot_factor_similarity(
  similarity_result = sim_BBMF,
  x_label = "BBMF Estimated Factors",
  y_label = "True Factors",
  title = "Factor Similarity"
)
```

### Align $H$ and $W$

The permutation obtained from $H$ is applied to the rows of $H$ and the corresponding columns of $W$.

```{r}
#| label: align-WH

perm_BBMF <- sim_BBMF$permutation

H_max_aligned <- H_max[
  perm_BBMF,
  ,
  drop = FALSE
]

W_max_aligned <- W_max[
  ,
  perm_BBMF,
  drop = FALSE
]

rownames(H_max_aligned) <- factor_names
colnames(H_max_aligned) <- arms

rownames(W_max_aligned) <- sample_ids
colnames(W_max_aligned) <- factor_names
```

## Compare the factor-feature matrices

The true and estimated $H$ matrices are displayed separately from the sample-factor matrices.

```{r}
#| label: compare-H
#| fig-width: 11
#| fig-height: 6

H_max_df <- reshape2::melt(
  H_max_aligned
)

H_max_df$Var1 <- factor(
  H_max_df$Var1,
  levels = factor_names
)

H_BBMF_graph <- matrixPlot(
  H_max_df,
  Var2,
  Var1,
  value,
  title = "BBMF Estimate of H"
)

patchwork::wrap_plots(
  pH,
  H_BBMF_graph,
  ncol = 1
)
```

This comparison provides a direct view of the latent feature patterns recovered by BBMF.

## Compare the sample-factor matrices

The corresponding $W$ matrices show which latent factors are active for each sample.

```{r}
#| label: compare-W
#| fig-width: 10
#| fig-height: 8

W_max_df <- reshape2::melt(
  W_max_aligned
)

W_max_df$Var2 <- factor(
  W_max_df$Var2,
  levels = factor_names
)

W_BBMF_graph <- matrixPlot(
  W_max_df,
  Var2,
  Var1,
  value,
  title = "BBMF Estimate of W"
)

patchwork::wrap_plots(
  pW,
  W_BBMF_graph,
  ncol = 2
)
```

## Aggregate Factor Alignment Similarity

The Aggregate Factor Alignment Similarity (AFAS) summarizes the similarity between a reference factorization and an estimated factorization after optimal factor matching.

For posterior sample $H^{(m)}$,

$$\operatorname{AFAS}^{(m)}
=
\frac{1}{R}
\sum_{r=1}^{R}
J
\left(
H^{\mathrm{true}}_{r\cdot},
H^{(m)}_{\sigma_m^*(r)\cdot}
\right),$$

where $J(\cdot,\cdot)$ denotes Jaccard similarity and $\sigma_m^*$ denotes the optimal factor assignment.

Values closer to one indicate greater similarity between the estimated and reference latent structures.

```{r}
#| label: compute-afas

afas_by_chain <- lapply(
  fit_BBMF,
  function(chain_fit) {

    get_mean_sim_per_chain(
      H_true = H_true,
      ff = chain_fit$samples_H
    )
  }
)

afas_matrix <- do.call(
  cbind,
  afas_by_chain
)

afas_matrix <- afas_matrix[
  -1,
  ,
  drop = FALSE
]

colnames(afas_matrix) <- paste0(
  "Chain_",
  seq_len(ncol(afas_matrix))
)

round(
  apply(
    afas_matrix,
    2,
    mean
  ),
  3
)
```

AFAS for the selected posterior solution is

```{r}
#| label: afas-selected

AFAS_H_max <- factor_similarity(
  H_ref = H_true,
  H_est = H_max
)$mean_similarity

round(
  AFAS_H_max,
  3
)
```

## Reconstruction performance

The function `bmf_metrics()` compares a Boolean reconstruction with a reference binary matrix.

Here we compare reconstruction from Asso and sparse BBMF with the true simulated matrix.

```{r}
#| label: reconstruction-metrics

X_hat_list <- list(
  Asso = X_asso,
  Sparse_BBMF = X_tilde_BBMF
)

Metrics <- do.call(
  rbind,
  lapply(
    names(X_hat_list),
    function(method) {

      metrics <- bmf_metrics(
        X = X_true,
        X_hat = X_hat_list[[method]]
      )

      data.frame(
        Method = method,
        t(metrics),
        row.names = NULL,
        check.names = FALSE
      )
    }
  )
)

numeric_cols <- vapply(
  Metrics,
  is.numeric,
  logical(1)
)

Metrics[numeric_cols] <- round(
  Metrics[numeric_cols],
  3
)

Metrics
```

The reported measures include specificity, F1-score, Matthews correlation coefficient, and reconstruction error rate.

## Posterior reconstruction probabilities

BBMF retains posterior samples of $W$ and $H$. These samples can be used to reconstruct $X$ across the posterior distribution.

First, the posterior arrays are extracted from each chain.

```{r}
#| label: posterior-X-samples
#| results: hide

Ws <- extract_samples_bbmf(
  model_object = fit_BBMF,
  sample_name = "samples_W"
)

Hs <- extract_samples_bbmf(
  model_object = fit_BBMF,
  sample_name = "samples_H"
)
```

The Boolean reconstruction is then calculated for each posterior sample.

```{r}
#| label: reconstruct-X-posterior
#| results: hide

X_reconstructed <- lapply(
  seq_along(Ws),
  function(chain_id) {

    reconstruct_X_one_chain_bbmf(
      W_samples = Ws[[chain_id]],
      H_samples = Hs[[chain_id]]
    )
  }
)

names(X_reconstructed) <- names(fit_BBMF)
```

Posterior reconstruction probabilities and uncertainty can then be summarized.

```{r}
#| label: posterior-X-summary

X_summary <- summarize_X_reconstruction_bbmf(
  reconstructed_chains = X_reconstructed,
  X_truth = X_true
)

X_probability <- X_summary$probability

X_uncertainty <- X_summary$uncertainty
```

For entry $(k,g)$, the posterior reconstruction probability is

$$\widehat{P}_{kg}
=
\frac{1}{M}
\sum_{m=1}^{M}
\widetilde{X}_{kg}^{(m)}.$$

A scaled measure of posterior uncertainty is

$$U_{kg}
=
4\widehat{P}_{kg}
\left(
1-\widehat{P}_{kg}
\right).$$

Values close to zero indicate high posterior agreement, while values close to one indicate greater uncertainty.

### Visualize posterior reconstruction probabilities

```{r}
#| label: posterior-X-probability
#| fig-width: 11
#| fig-height: 6

X_long <- make_X_long_data_bbmf(
  probability = X_probability,
  uncertainty = X_uncertainty,
  X_truth = X_true
)

plot_X_posterior_bbmf(
  long_data = X_long$probability,
  type = "probability",
  title = "Posterior Reconstruction Probability"
)
```

### Visualize posterior reconstruction uncertainty

```{r}
#| label: posterior-X-uncertainty
#| fig-width: 11
#| fig-height: 6

plot_X_posterior_bbmf(
  long_data = X_long$uncertainty,
  type = "uncertainty",
  title = "Posterior Reconstruction Uncertainty"
)
```

## Posterior inclusion probabilities for $H$

Posterior samples of $H$ cannot be averaged directly because factor labels may change across posterior samples.

The package therefore provides `align_H_samples_bbmf()` for aligning posterior samples to a reference factorization.

In this simulated example, the true $H$ matrix is used as the reference.

```{r}
#| label: posterior-H-alignment
#| results: hide

Hs_aligned <- lapply(
  Hs,
  function(H_chain) {

    align_H_samples_bbmf(
      H_ref = H_true,
      H_samples = H_chain
    )
  }
)

names(Hs_aligned) <- names(Hs)
```

The aligned posterior samples can then be summarized using `summarize_H_posterior_bbmf()`.

```{r}
#| label: posterior-H-summary

H_summary <- summarize_H_posterior_bbmf(
  H_chains = Hs_aligned,
  H_ref = H_true
)

H_probability <- H_summary$probability

H_uncertainty <- H_summary$uncertainty
```

For factor $r$ and feature $g$, the posterior inclusion probability is

$$\widehat{P}_{rg}^{H}
=
\frac{1}{M}
\sum_{m=1}^{M}
H_{rg}^{(m)}.$$

### Posterior inclusion probability for $H$

```{r}
#| label: posterior-H-probability
#| fig-width: 11
#| fig-height: 4

plot_H_posterior_bbmf(
  matrix_summary = H_probability,
  type = "probability",
  title = "Posterior Inclusion Probabilities for H"
)
```

### Posterior uncertainty for $H$

```{r}
#| label: posterior-H-uncertainty
#| fig-width: 11
#| fig-height: 4

plot_H_posterior_bbmf(
  matrix_summary = H_uncertainty,
  type = "uncertainty",
  title = "Posterior Uncertainty for H"
)
```

The posterior inclusion probabilities summarize how frequently individual factor-feature associations are present across posterior samples, while the uncertainty plot highlights associations for which the posterior distribution is less certain.

## Main BBMF functions

The following functions illustrate the main capabilities of the package.

| Function | Purpose |
|----|----|
| `bool_prod()` | Compute Boolean matrix products |
| `noise_bool()` | Introduce random noise into binary matrices |
| `Asso_BBMF()` | Obtain an Asso Boolean factorization using `rBMF` |
| `run_sparseBBMF()` | Fit a single sparse BBMF chain |
| `run_multiple_sparseBBMF()` | Fit multiple sparse BBMF chains |
| `diagnosBBMF()` | Compute MCMC diagnostic quantities |
| `group_diagnost_List()` | Organize diagnostics across chains |
| `matrixPlot()` | Visualize binary matrices |
| `jaccard_rows()` | Compute row-wise Jaccard similarities |
| `jaccard_cols()` | Compute column-wise Jaccard similarities |
| `factor_alignment_hungarian()` | Align latent factors using optimal assignment |
| `factor_similarity()` | Assess similarity between two factor matrices |
| `plot_factor_similarity()` | Visualize factor similarity |
| `get_mean_sim_per_chain()` | Calculate AFAS across posterior samples |
| `bmf_metrics()` | Evaluate Boolean reconstruction performance |
| `extract_samples_bbmf()` | Extract posterior samples from BBMF chains |
| `reconstruct_X_one_chain_bbmf()` | Reconstruct $X$ from posterior samples |
| `summarize_X_reconstruction_bbmf()` | Summarize posterior reconstruction of $X$ |
| `plot_X_posterior_bbmf()` | Visualize posterior summaries for $X$ |
| `align_H_samples_bbmf()` | Align posterior samples of $H$ |
| `summarize_H_posterior_bbmf()` | Calculate posterior summaries for $H$ |
| `plot_H_posterior_bbmf()` | Visualize posterior summaries for $H$ |

## Typical workflow

A typical analysis using `BBMF` follows the general workflow

``` text
Observed binary matrix X
          |
          v
Initial Boolean factorization
          |
          v
Sparse BBMF
          |
          +----> MCMC diagnostics
          |
          +----> Posterior samples of W and H
          |          |
          |          +----> factor alignment
          |          +----> factor similarity / AFAS
          |          +----> posterior inclusion probabilities
          |          +----> posterior uncertainty
          |
          +----> Boolean reconstruction of X
                     |
                     +----> reconstruction metrics
                     +----> posterior reconstruction probabilities
                     +----> posterior reconstruction uncertainty
```

The `BBMF` package therefore provides a workflow for estimating interpretable Boolean latent patterns while retaining the posterior information needed to evaluate factor stability, reconstruction performance, and uncertainty.

## Acknowledgements

The `BBMF` package implements the Bayesian Boolean matrix factorization methodology described by Wagala, Samur, and Parmigiani [@wagala2026bbmf].

The package uses functionality from several existing R packages. In particular, `rBMF` provides the Asso Boolean matrix factorization used for initialization and comparison. Packages including `ggplot2`, `reshape2`, and `patchwork` support visualization and data preparation, while additional R packages provide computational utilities used internally by `BBMF`.

## References

::: {#refs}
:::
